### 2017-12-11 submission/manuscript.Rmd
update abstract with more direct sentences

**Results.** Fecal samples had a significant decrease from control to adenoma to carcinoma for both Shannon diversity and evenness (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) after correcting for study effect and variable region sequenced. Only evenness for adenoma (P-value `r ifelse(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) resulted in a slightly increased relative risk while lower Shannon diversity and evenness in fecal samples resulted in a significant increase in relative risk for carcinoma (P-value `r ifelse(min(filter(alpha_RR_crc_stool, !(measure %in% "sobs"))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). Previously associated colorectal cancer genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) followed a similar pattern with a significantly increased relative risk by their presence for carcinoma (P-value `r ifelse(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) but not adenoma (P-value `r ifelse(min(filter(adn_select_RR_stool, !(measure %in% c("all_four", "total_four", "Porphyromonas")))[, "pvalue"]) > 0.05, invisible(">"), invisible("check tables"))` 0.05) with the exception of *Porphyromonas* (P-value `r ifelse(min(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). Using the whole community versus only CRC associated genera to build a prediction model resulted in higher classification success based on Area Under the Curve (AUC) for both adenoma and carcinoma using fecal and tissue samples. The most important OTUs for these models consistently belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* across studies. For the included studies, most were adequately powered for large effect size differences. This may be more amenable for carcinoma than for adenoma microbiota research due to the smaller community level changes observed in our results.

**Conclusions.** This data provides support for the importance of the bacteral community to both adenoma and carcinoma genesis. The evidence collected within this study on the associations between the microbiota and CRC are much stronger for carcinoma then adenoma. A strong reason for this may be in part due to the low power to detect more subtle changes in the majority of studies that have been performed to date.

>>>>

**Results.** Fecal samples had a significant decrease from control to adenoma to carcinoma for both Shannon diversity and evenness after correcting for study effect and variable region sequenced (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). This reduction in evenness resulted in small increases in relative risk for adenoma (P-value = `r format(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]), digits = 2)`) and carcinoma (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2)`) while the reduction in Shannon diversity only resulted in an increased relative risk for carcinoma (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2)`). Previously associated colorectal cancer genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) followed a similar pattern, with their presence significantly increasing the relative risk for carcinoma (P-value `r ifelse(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) but not adenoma (P-value `r ifelse(min(filter(adn_select_RR_stool, !(measure %in% c("all_four", "total_four", "Porphyromonas")))[, "pvalue"]) > 0.05, invisible(">"), invisible("check tables"))` 0.05) with the exception of *Porphyromonas* (P-value = `r format(min(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"]), digits = 2)`). Using the whole community versus only CRC associated genera to build a prediction model resulted in higher classification success based on Area Under the Curve (AUC) for both adenoma and carcinoma using fecal and tissue samples. The most important OTUs for these models consistently belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* across studies. Overall, there were less associations between the microbiota and adenoma versus carcinoma and one reason why this may be is that most studies were only adequatly powered for large effect sizes.

**Conclusions.** This data provides support for the importance of the bacteral community to both adenoma and carcinoma genesis. The evidence collected within this study on the role of the microbiota in CRC pathogenesis shows stronger associations between carcinoma then adenoma. A strong reason for this may be in part due to the low power to detect more subtle changes in the majority of studies that have been performed to date.

### 2017-12-11 submission/manuscript.tex
update abstract with more direct sentences

\textbf{Results.} Fecal samples had a significant decrease from control
to adenoma to carcinoma for both Shannon diversity and evenness (P-value
\textless{} 0.05) after correcting for study effect and variable region
sequenced. Only evenness for adenoma (P-value \textless{} 0.05) resulted
in a slightly increased relative risk while lower Shannon diversity and
evenness in fecal samples resulted in a significant increase in relative
risk for carcinoma (P-value \textless{} 0.05). Previously associated
colorectal cancer genera (\emph{Fusobacterium}, \emph{Parvimonas},
\emph{Peptostreptococcus}, or \emph{Porphyromonas}) followed a similar
pattern with a significantly increased relative risk by their presence
for carcinoma (P-value \textless{} 0.05) but not adenoma (P-value
\textgreater{} 0.05) with the exception of \emph{Porphyromonas} (P-value
\textless{} 0.05). Using the whole community versus only CRC associated
genera to build a prediction model resulted in higher classification
success based on Area Under the Curve (AUC) for both adenoma and
carcinoma using fecal and tissue samples. The most important OTUs for
these models consistently belonged to genera such as
\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} across
studies. For the included studies, most were adequately powered for
large effect size differences. This may be more amenable for carcinoma
than for adenoma microbiota research due to the smaller community level
changes observed in our results.

\textbf{Conclusions.} This data provides support for the importance of
the bacteral community to both adenoma and carcinoma genesis. The
evidence collected within this study on the associations between the
microbiota and CRC are much stronger for carcinoma then adenoma. A
strong reason for this may be in part due to the low power to detect
more subtle changes in the majority of studies that have been performed
to date.

>>>>

\textbf{Results.} Fecal samples had a significant decrease from control
to adenoma to carcinoma for both Shannon diversity and evenness after
correcting for study effect and variable region sequenced (P-value
\textless{} 0.05). This reduction in evenness resulted in small
increases in relative risk for adenoma (P-value = 0.032) and carcinoma
(P-value = 0.00034) while the reduction in Shannon diversity only
resulted in an increased relative risk for carcinoma (P-value = 0.0047).
Previously associated colorectal cancer genera (\emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas})
followed a similar pattern, with their presence significantly increasing
the relative risk for carcinoma (P-value \textless{} 0.05) but not
adenoma (P-value \textgreater{} 0.05) with the exception of
\emph{Porphyromonas} (P-value = 0.023). Using the whole community versus
only CRC associated genera to build a prediction model resulted in
higher classification success based on Area Under the Curve (AUC) for
both adenoma and carcinoma using fecal and tissue samples. The most
important OTUs for these models consistently belonged to genera such as
\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} across
studies. Overall, there were less associations between the microbiota
and adenoma versus carcinoma and one reason why this may be is that most
studies were only adequatly powered for large effect sizes.

\textbf{Conclusions.} This data provides support for the importance of
the bacteral community to both adenoma and carcinoma genesis. The
evidence collected within this study on the role of the microbiota in
CRC pathogenesis shows stronger associations between carcinoma then
adenoma. A strong reason for this may be in part due to the low power to
detect more subtle changes in the majority of studies that have been
performed to date.
